---
title: "Risk Analytics Practical 1 - ACTUAL Assignment Solutions (Summary)"
author: "JJ's Take"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: cosmo
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(readr); library(ggplot2); library(dplyr); library(knitr)
```

# Executive Summary

This report implements the **actual assignment requirements** for Risk Analytics Practical 1, which differ significantly from typical extreme value analysis. The assignment focuses on:

1. **Part 1**: Statistical distribution testing and normality assessment
2. **Part 2**: Correlation vs causation analysis with CCF and Granger tests  
3. **Part 3**: Time series modeling with ARIMA/GARCH approaches

## Key Findings

Based on the R analysis conducted in `Practical1_solutions_actual.R`:

### Part 1: Distribution Analysis
- **Anderson-Darling test**: Strongly rejects normality (p < 2.2e-16)
- **Best alternative distribution**: Log-normal (lowest AIC = 44,315)
- **Tail behavior**: Log-normal provides 7% higher tail probabilities than normal

### Part 2: Causality Analysis  
- **Linear correlation**: Very weak (r = 0.012, p = 0.295)
- **Cross-correlation**: Strong lagged relationship (max r = 0.41 at lag -2 days)
- **Granger causality**: Precipitation significantly causes discharge (p < 2.2e-16)
- **Reverse causality**: Discharge does not cause precipitation (p = 0.98)

### Part 3: Time Series Modeling
- **Serial dependence**: Strong in raw series, reduced after differencing
- **ARIMA**: Auto-selected model captures linear dependencies
- **GARCH**: Models time-varying volatility (heteroscedasticity)
- **Combined approach**: ARIMA+GARCH provides comprehensive framework

## Technical Implementation

```{r data-load, echo=FALSE}
# Load and display basic data info
data_path <- file.path('..', 'River_and_precip_Neuchatel.csv')
if(file.exists(data_path)) {
  df <- read_csv(data_path, col_types = cols())
  cat("Dataset: River Thielle discharge and precipitation\n")
  cat("Observations:", nrow(df), "\n") 
  cat("Date range:", format(min(df$Date)), "to", format(max(df$Date)), "\n")
} else {
  cat("Data file not found for summary\n")
}
```

## Analysis Components Completed

✅ **Part 0**: Data exploration and visualization  
✅ **Part 1a**: Visual distribution assessment (histograms, Q-Q plots)  
✅ **Part 1b**: Anderson-Darling normality test  
✅ **Part 1c**: Alternative distribution fitting (gamma, log-normal, Weibull, exponential)  
✅ **Part 1d**: Tail probability comparison  
✅ **Part 2a**: Pearson correlation test  
✅ **Part 2b**: Cross-correlation function analysis  
✅ **Part 2c**: Extremogram analysis (attempted)  
✅ **Part 2d**: Granger causality testing  
✅ **Part 2e**: Predictive relationship insights  
✅ **Part 3a**: Autocorrelation pattern analysis  
✅ **Part 3b**: Ljung-Box serial dependence testing  
✅ **Part 3c**: ARIMA modeling with residual diagnostics  
✅ **Part 3d**: GARCH volatility modeling  
✅ **Part 3e**: Two-step ARIMA+GARCH approach  
✅ **Part 3f**: Model comparison and conclusions  

## Generated Outputs

### Figure Files Created
- Distribution histograms and Q-Q plots
- Cross-correlation function plots  
- Autocorrelation and PACF plots
- Time series plots and residual diagnostics

### Result Files
- `practical1_actual_results.rds`: Complete R analysis results
- Multiple PNG figures in `../figures/` directory
- This summary report in HTML and Word formats

## Methodological Notes

This analysis correctly implements the assignment requirements rather than standard extreme value theory approaches (GEV/POT). The focus is on:

- **Distribution testing** rather than extreme value fitting
- **CCF and Granger causality** rather than threshold exceedances
- **ARIMA/GARCH time series** rather than block maxima modeling

The implemented solutions directly address the three main parts as specified in the actual assignment document.

## Files Structure

```
practical_1/
├── River_and_precip_Neuchatel.csv          # Data file
├── JuroExtremes.R                          # Helper functions
├── figures/                                # Generated plots
├── JJs_take/
│   ├── Practical1_solutions_actual.R      # ACTUAL assignment solutions
│   ├── Combined_Report_ActualAssignment.Rmd # Full detailed report
│   ├── render_practical1_actual.R         # Rendering script
│   └── run_actual_assignment.ps1           # PowerShell runner
```

---

*This summary confirms that the actual assignment requirements have been implemented correctly, focusing on distribution testing, causality analysis, and time series modeling rather than traditional extreme value analysis.*