---
title: "Practical 1 — River discharge & precipitation extremes (JJ's take)"
author: ""
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: pdflatex
    toc: true
    number_sections: false
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4)
library(knitr)
```

# NOTE
This is a copy of the working report placed in `practical_1/JJs_take/` for teammates. The data and helper functions are shared in the parent `practical_1/` directory: keep `River_and_precip_Neuchatel.csv` and `JuroExtremes.R` in `practical_1/`.

# How to run
Open PowerShell and from the project root run the runner script:

```powershell
.\practical_1\JJs_take\run_from_JJs_take.ps1
```

Or run the solution script directly:

```powershell
& 'C:\Program Files\R\R-4.5.1\bin\Rscript.exe' -e "setwd('C:/Users/jskif/OneDrive - epfl.ch/Documents/SMT/RA/risk_analytics'); Sys.setenv(RSTUDIO_PANDOC='C:/Program Files/Pandoc'); source('practical_1/JJs_take/Practical1_solutions.R')"
```

# Working copies
This folder contains copies of the solution script and the report for JJ's edits. Do not forget to save final outputs (figures/RDS) in the parent `practical_1/` figures folder to keep a single canonical location for outputs.

```{r echo=FALSE}
cat('See README.md in this folder for more details')
```
